---
title: "Gov 50 Final Project"
author: "Talulah Ratcliffe"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

## Introduction
Is the inclusion of 'Diversity, Equity and Inclusion' within K-12 School District mission statements affected by party, race and income? In this study, I plan to examine the extent to which socioeconomic and partisan factors of a given school district in the US, influences whether or not values of diversity, equity and inclusion are included in their mission statements. 

As seen through the Affirmative Action case this year, conservative ideology tends to value race-blind and, more generally, identity-blind policies and values. Policies (with a school Mission statement replicating the effect of a legal imposition) which attempt to fundamentally interfere with objective principles which equally apply to all, are usually rejected by Republicans, as well as privileged racial and economic groups. 

As such, I hypothesize that there will not only be a correlation between party and race/income, but that Republican districts, particularly those with higher proportions of white, upper-class students, will be less likely to include Diversity, Equity and Inclusion within their mission statements. 


## Data
My representative sample is comprised of 1314 mission statements from regular public school districts, whose mission statements were published online as of November 2022. The schools were chosen from a comprehensive list maintained by the National Center for Education Statistics. My unit of analysis is a Mission Statement. The explanatory variable of interest is whether values of diversity, equity or inclusion are included. The variable is coded = 1 for included, and = 0 for excluded. There are several key words that will indicate these values (closing gap, culture, diversity, equity, inclusion, nationality, race). My main outcome variable is the party of the district - Republican or Democrat. This is reflected through whether the majority of residents voted for Donald Trump or Joe Biden in the 2020 presidential election, and is calculated using 2020 General Election Results Disaggregated to 2020 Census Blocks Datasets available through the Redistricting Data Hub. Population data includes median income and share of white residents, and was accessed from the 2015-2019 American Community Survey estimates published by the Census Bureau. My research design is cross-sectional, as a single representative sample was collected from one time point.  

If I observe less inclusion of diversity, inclusion and equity values in Mission Statements within Republican, upper-class and predominately white districts, this would provide support for my hypothesis. On the other hand, if I observe less inclusion of diversity, inclusion and equity values in Mission Statements within Democratic, lower-class and predominately non-white school districts, this would provide evidence against my hypothesis. 

```{r loading-data, echo=FALSE}
library(tidyverse)
diversity_data <- read.csv("data/district_tabular_data.csv")
```


```{r cleaning-data, echo=FALSE}
diversity_data <- diversity_data |>
  select(topic_diversity_equity_and_inclusion, majority_vote, white_share_category, hh_median_income_category)
```

```{r initial-count, echo=FALSE}
ggplot(diversity_data, aes(x = topic_diversity_equity_and_inclusion)) +
  geom_bar(stat = "count", fill = "lightblue") +
  geom_text(
    aes(label = after_stat(count)),
    stat = "count",
    vjust = -0.5,
    size = 3       
  ) +
  geom_text(
    aes(label = ifelse(topic_diversity_equity_and_inclusion == 1, "Included", "Not Included"), y = 0),
    vjust = -1.5,
    hjust = 0.5,
    size = 3
  ) +
  labs(
    title = "Distribution of DEI Topics in Mission Statements",
    x = "DEI Topic Inclusion",
    y = "Count",
  ) +
  theme_minimal() 
```
As seen, there were 708 mission statements which did not include DEI topics, and 606 mission statements that did. 

## Results

```{r only-party, echo=FALSE}
div_count <- diversity_data |>
  group_by(majority_vote) |>
  filter(!is.na(majority_vote), topic_diversity_equity_and_inclusion == 1) |>
  count()

ggplot(div_count, aes(x = majority_vote, y = n)) +
  geom_bar(stat = "identity", fill = c("democratic" = "lightblue", "republican" = "lightcoral")) +
  labs(
    title = "Diversity, Equity and Inclusion Topic by Party",
    x = "Party Affiliation",
    y = "Number of Mentions"
  ) +
  scale_x_discrete(labels = c("democratic" = "Democratic", "republican" = "Republican")) +
  theme_minimal() 
```

This graph shows a comparison of democratic and republican inclusions of diversity, equity or inclusion topics within their mission statements. It is the beginning of an exploration between party affiliation and an educational focus on these topics, and suggests that the district being democratic increases a school's tendency to include them. However, I also wish to investigate if there are any connections between party, race and income level, as my hypothesis grouped these socioeconomic factors with party affiliation. 


```{r income-party-table, echo=FALSE}
income_table <- diversity_data |>
  mutate(Income = case_when(
  hh_median_income_category %in% c("low_median_income") ~ "Low Median Income",
  hh_median_income_category %in% c("medium_median_income") ~ "Medium Median Income",
  hh_median_income_category %in% c("high_median_income") ~ "High Median Income"),
    Party = case_when(
      majority_vote == "republican" ~ "Republican",
      majority_vote == "democratic" ~ "Democratic",
      NA ~ "Other"
    )) |>
  group_by(Party, Income) |>
  summarize(n = n()) |>
  mutate(prop = n / sum(n)) |>
  select(-n) |>
  pivot_wider(
    names_from = Party,
    values_from = prop
  ) |>
select(Income, Democratic, Republican)
knitr::kable(income_table)
```
```{r income-party-graph, echo=FALSE}
party_income <- income_table |>
  pivot_longer(cols = c("Democratic", "Republican"), names_to = "Party", values_to = "Proportion")

ggplot(party_income, aes(x = Income, y = Proportion, fill = Party)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Proportion of Democratic and Republican Voters by Income",
    x = "Income Group",
    y = "Proportion"
  ) +
  scale_fill_manual(values = c("Democratic" = "lightblue", "Republican" = "lightcoral")) +
  theme_minimal()
```


This proportion table, and accompanying graph, examines how party differs by income levels, in an attempt to see if there is a connection between the two. It does so by showing the proportion of each party within each income group. Both table and graph show that both the high and low median income districts have a relatively balanced distribution between Democrats and Republicans. However, the proportion of democrats is higher than republicans in both high and low income categories, with the difference being slightly higher in the high income group. However, the medium median income group shows the largest disparity, with most being Republican.

These findings are somewhat aligned with my hypothesis. I grouped higher median income with Republican districts, but that is more true for medium income / middle-class districts. However, my assumed grouping of low median income groups with democratic areas is plausible based on this analysis. 


```{r party-race, echo=FALSE}
diversity_data |>
  drop_na(majority_vote, white_share_category) |>
  count(majority_vote, white_share_category) |>
ggplot(mapping = aes(x = white_share_category, y = n, fill = majority_vote)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Correlation between Party and Race",
       x = "Predominant Race of District Residents",
       y = "Frequency",
       fill = "Party Affiliation") +
  scale_fill_manual(values = c("democratic" = "lightblue", "republican" = "lightcoral"),
                    name = "Party Affiliation",
                    labels = c("democratic" = "Democratic", "republican" = "Republican")) +
  scale_x_discrete(labels = c("low_share" = "Predominately Non-White Population", "high_share" = "Predominately White Population")) +
  theme_minimal()
```
This graph investigates if there is a connection between party affiliation and race within a given school district. There are more democratic areas that are predominately non-white, than there are republican areas. In comparison, there are far more republican areas when the population is predominately white, and the difference is even more pronounced. Notably, there are less overall values in the predominately white population category, which could suggest an imbalance in the data. 

The implied connection between race and party, wherein districts with a higher proportion of white residents tend to be republican, and districts with a lower proportion of white residents tend to be democratic, aligns with my hypothesis which grouped these two variables together.  


```{r party-race-income, echo=FALSE}
combined_table <- diversity_data %>%
  mutate(
    Income = case_when(
      hh_median_income_category %in% c("low_median_income") ~ "Low Median Income",
      hh_median_income_category %in% c("medium_median_income") ~ "Medium Median Income",
      hh_median_income_category %in% c("high_median_income") ~ "High Median Income"
    ),
    Party = case_when(
      majority_vote == "republican" ~ "Republican",
      majority_vote == "democratic" ~ "Democratic",
      NA ~ "Other"
    )
  ) %>%
  group_by(white_share_category, Party, Income) %>%
  summarize(prop = n() / nrow(.))

filtered_data <- combined_table %>%
  filter(Party %in% c("Democratic", "Republican"))

ggplot(filtered_data, aes(x = Income, y = prop, fill = Party)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Proportions of Democratic and Republican Voters by Income and Race",
    x = "Income Category",
    y = "Proportion"
  ) +
  scale_fill_manual(values = c("Democratic" = "lightblue", "Republican" = "lightcoral")) +
  facet_grid(white_share_category ~ ., scales = "free_y", space = "free_y", switch = "y", labeller = labeller(white_share_category = c("low_share" = "Predominately Non-White Population", "high_share" = "Predominately White Population"))) +
  theme_minimal() +
  theme(strip.text = element_text(size = 8))
```
This graph groups all 3 variables: race, party and income group, to assess if there are any consistent demographic links. The results indicate the following: High income, predominately non-white districts are mostly democratic. High income, predominately white districts are mostly republican. Low income, predominately non-white districts are mostly democratic. Low income, predominately white districts are mostly republican. Medium income, predominately non-white districts are mostly democratic. Medium income, predominately white districts are mostly republican. This indicates that across income groups, race is the predominate factor in predicting party affiliation. However, the medium income group does produce a much higher proportion of republicans. These findings align with my hypothesis groupings, which grouped non-white, low income and democratic districts as one distinct group, and white, high income, republic districts as another. 

Now, I want to connect these unique groups with whether or not they included the topics of diversity, equity or inclusion in their mission statements. In particular, my hypothesis suggested that high income, Republican districts, with predominately white populations, would be the least likely to include Diversity, Equity and Inclusion within their mission statement. 


```{r unique-group-DEI, echo=FALSE}
count_by_group <- diversity_data |>
  na.omit() |>
  group_by(hh_median_income_category, white_share_category, majority_vote) |>
  summarize(count = sum(topic_diversity_equity_and_inclusion == 1)) |>
  mutate(
    hh_median_income_category = case_when(
      hh_median_income_category == "low_median_income" ~ "Low Income",
      hh_median_income_category == "medium_median_income" ~ "Medium Income",
      hh_median_income_category == "high_median_income" ~ "High Income",
      TRUE ~ hh_median_income_category
    ),
    white_share_category = case_when(
      white_share_category == "low_share" ~ "Mostly White",
      white_share_category == "high_share" ~ "Mostly Non-White",
      TRUE ~ white_share_category
    ),
    majority_vote = case_when(
      majority_vote == "republican" ~ "Republican",
      majority_vote == "democratic" ~ "Democratic",
      TRUE ~ majority_vote
    )
  ) |>
  arrange(desc(count))

knitr::kable(count_by_group, col.names = c("Income Group", "Predominate Race", "Party", "Count"),
             caption = "**Summary of DEI Topics Mentions by Income Group, Predominate Race, and Party**")
```
These findings organize the unique socioeconomic/partisan groups from most inclusions of DEI topics, to least. It reveals that the group with the most DEI topic mentions was mostly white, democratic, medium income districts, closely followed by mostly white, high income, democratic districts. The group with the least inclusions in their mission statements were low income and medium income, mostly non-white, democratic districts. Interestingly, all republican districts fell in the middle of the distribution, regardless of race breakdown. These results challenge the initial hypothesis, indicating that assumptions about Republican districts, high-income districts, and mostly white districts being less likely to include DEI topics may not hold true in this dataset. The unexpected patterns observed underscore the complexity of factors influencing the inclusion of DEI topics in mission statements. 

To explore the nuanced impact of party affiliation on the inclusion of DEI topics in mission statements while accounting for race and income, a regression analysis provides valuable insights. 

```{r nice-regression, echo=FALSE}
library(gt)
library(broom)
party_reg <- lm(topic_diversity_equity_and_inclusion ~ majority_vote + white_share_category + hh_median_income_category, data = diversity_data)
tidy_regression <- tidy(party_reg)

tidy_regression |>
  gt() |>
  tab_header(
    title = "Regression Results",
    subtitle = "Predicting Inclusion of DEI Topics"
  ) |>
  fmt_number(
    columns = vars(estimate, std.error, statistic, p.value),
    decimals = 3
  )
```
This regression has an intercept of 0.495. This means that when all other variables (party, race, and income) are at their reference levels, which would represent mostly white residents, a high median income, and a Democratic district, the estimated value of the inclusion of DEI topics is approximately 0.495. This value is closer to 0 (not including) than 1 (including), suggesting that, on average, areas with predominantly white residents, high median income, and a Democratic majority are less likely to include Diversity, Equity, and Inclusion (DEI) topics in their mission statements compared to the overall average. The p-value is below 0.05, which indicates that this is statistically significant, and that there is evidence to reject the null hypothesis that the intercept is zero (and thus wouldn't be meaningful). 

My main coefficient of interest is **party affiliation**. The estimate for majority_voterepublican is -0.236. This means that when all other variables are held constant, the estimated change in the inclusion of DEI topics with a Republican, rather than Democratic, majority is -0.2359. In other words, on average, areas with a Republican majority are 23.59% less likely to include DEI topics compared to a district with a Democratic majority, holding all other factors constant. Its p value is lower than 0.05, which means it is also statistically significant, and shows that the effect of having a Republican district is most likely meaningful; we can conclude it has a causal effect on the inclusion of DEI topics.

The estimate for white_share_categorylow_share is 0.166. This means that when all other variables are held constant, the estimated change in the inclusion of DEI topics with a low share of white residents rather than a high share, is approximately 0.1664. In other words, on average, areas with mostly non-white residents are 16.64% more likely to include DEI topics compared to a district with mostly white residents, holding all other factors constant. The low p value similarly indicates it is a meaningful causal factor and statistically significant.


The estimate for hh_median_income_categorylow_median_income is -0.080. This means that when all other variables are held constant, the estimated change in the inclusion of DEI topics with a low median income (compared to a high median income), is approximately -0.080. In other words, on average, low income districts are 8% less likely to include DEI topics than high income districts, holding all other factors constant. The low p value similarly indicates it is a meaningful causal factor and statistically significant. 

The estimate for hh_median_income_categorymedium_median_income is -0.008. This means that when all other variables are held constant, the estimated change in the inclusion of DEI topics with a medium median income (compared to a high median income), is approximately -0.008. In other words, on average, medium income districts are 0.8% less likely to include DEI topics than high income districts, holding all other factors constant. However, the p value for the median income group is greater than 0.05, suggesting there is insufficient evidence to reject the null hypothesis that the effect of having a medium median income is zero; it is unclear whether being from a medium income district is a causal effect on whether a school includes DEI topics.


## Conclusion 
Overall, this report has investigated whether party, race and income of a given district, affect a school's likeliness to include topics of Diversity, Equity or Inclusion in their mission statement. Before connecting either socioeconomic or partisan variables to the inclusion of DEI topics, associations between party, race and income were explored. It was found that there was a general connection between income and party, where both high and low income districts were more likely to be democratic than republican, whereas medium income districts were mostly republican. In terms of race, districts with mostly white residents were more likely to be republican, and districts with mostly non-white residents were most likely democratic. This correlation between race and party was proven to be true even accounting for income groups. Although the group with the least inclusions of DEI topics was low income, mostly non-white, democratic districts, which appeared to disprove every aspect of my hypothesis, the regression analysis provided more nuance. Schools in republican districts were significantly less likely to include DEI topics, and schools in districts with mostly non-white residents were more likely than mostly white districts to include DEI topics. However, the regression showed that high income districts were the most likely income group to include DEI topics - both low and medium income districts were less likely. This was against my hypothesis, which suggested that high income districts would be the least likely to include DEI topics. 

There were several limitations of the analysis and threats to inference that could be improved with future research. 
Within my data manipulation, I constantly excluded NA values to produce cleaner visualizations and ensure the analysis was based on complete cases. However, this may have left the dataset unbalanced and reduced the sample size. Moreover, the sample may not be fully representative of all public school districts, potentially introducing selection bias. Similarly, these findings may not be generalizable outside of the U.S, where political polarization is more prevalent, and connections between socioeconomic factors have been affected by historical segregation policies. There may also be unobserved confounding variables which could influence both the inclusion of DEI topics, and party affiliation. This can also be applied to the limited temporal scope of the data, as changes in political or societal attitudes over time may confound the relationship between party affiliation and DEI inclusion. There are several recommendations to counter these limitations: increase the sample size to capture more diverse schools across the world, conduct further analysis on whether certain NA values are meaningfully connected to different socioeconomic/partisan variables, introduce more variables for analysis (such as the age of the school, the racial/class makeup of the school itself, predominate religion of the district etc.) and examine how these correlations may change over time. Moreover, the initial binary focus on only high/low income groups may have missed important conclusions that could be drawn from the medium income group. Further investigation into this may prove insightful. 
